---
title: "Why Chart?"
author: "Philippe Grosjean"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Why Chart?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5, out.width = "95%")
library(chart)
```

In this vignette, we explain some of the design choices behind the {chart} package, and why it can be useful for R users.

## Formula versus `aes()`

Since the formula interface is used extensively in R to specify models, it is a little bit strange that corresponding plots with {ggplot2} must be specified in a totally different way. The {chart} package allows to specify the same "aesthetics" through the familiar formula interface. For instance, here is a simple model:

```{r}
data(trees)
trees_lm <- lm(Volume ~ Girth, data = trees)
summary(trees_lm)
```
Our model is specified with `Volume ~ Girth`, which must be translated into `aes(x = Girth, y = Volume)` in {ggplot2}. A common pitfall for beginners is to invert X and Y here because they appear in a different order in both interfaces.

```{r}
library(ggplot2)
ggplot(trees, aes(x = Girth, y = Volume)) +
  geom_point() +
  geom_smooth(method = "lm", formula = 'y ~ x')
```


With {chart}, we can use the formula interface directly:

```{r}
library(chart)
chart(trees, Volume ~ Girth) +
  geom_point() +
  geom_smooth(method = "lm", formula = 'y ~ x')
```

Of course, it is not necessarily the **same** formula that is used for the model and the chart. The formula interface has different meaning, depending on the function that uses it. To specify more aesthetics in {chart} than just `x` and `y`, we use `%arg=%` terms. For instance, to specify the color of the points, we use `%col=%`:

```{r}
chart(trees, Volume ~ Girth %col=% Height) +
  geom_point()
```

## The generic function `chart()`

On the contrary to `ggplot()`, `chart()` is a generic. It means you do not have to specify a **data frame** for `data =`, but you can provide any object that has `chart()`, or `autoplot()` methods defined. There are not many methods defined in the {chart} package itself. Here is the method for **function** objects that create base R plots, converted into {chart} compatible objects:

```{r}
chart(function() {
  plot(Volume ~ Girth, data = trees, pch = 19, col = "black")
  grid(lty = "solid")
})
```

The {chart} package is designed as a basic framework to implement several methods for different R objects. It uses the "subsettable type" semantic, that is, its `type =` argument can define different plots for the same object class, and then, calling `chart$<type>(....)` produces the given type of plot for your object.

TODO: make a short example here.

See packages like [{modelit}](https://www.sciviews.org/modelit/) or [{exploreit}](https://www.sciviews.org/exploreit/) for several examples of such methods.

## `chart()` produces near publication-ready plots

Another aspect of `chart()` is that it is tailored to produce plots that are as close as possible to be publication-ready (but they. remain fully editable with {ggplot2} functions, of course). Hence, the default theme, and the labels of axes or legends are set up accordingly. If the data frame has **label** and **units** attributes defined for the variables, they are used to construct the labels of the axes and legends. Here is an example with the `trees` dataset:

```{r} 
trees <- data.io::labelise(trees, label = list(
    Volume = "Volume of timber",
    Girth  = "Diameter at 1.4m",
    Height = "Height"
  ), units = list(
    Volume = "cubic feet",
    Girth  = "inches",
    Height = "feet"))
chart(trees, Volume ~ Girth %col=% Height) +
  geom_point()
```
